---
title: "PhillyStat360 — Violation Tier Mapping"
subtitle: "VacancyGuide2026 Verification & Complete Building Tier Reference"
author: "Zhe Fang"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(scales)
library(knitr)
library(kableExtra)
```

# 0. Purpose

This document builds the **complete violation tier mapping** for the PhillyStat360 
building model. It replaces the earlier `violation_tier_mapping.xlsx` which contained
only a partial subset of codes.

**Why this matters:** The current VPI model scores violations on a 3-tier system 
(+1, +2, +3 points). To engineer violation-based features for our ML model, we need 
the exact code lists. Getting this wrong means our features won't align with the 
system we're trying to improve.

**What this document does:**

1. Extracts the **exact code lists** from VacancyGuide2026.pdf (pages 6–7)
2. Extracts the 54 "vacancy" codes from VIOLATION_DEFINITION.csv 
3. Cross-compares the two — shows where they overlap and where they don't
4. Builds the **correct, complete tier mapping** with evidence for every assignment
5. Defines the two-track query logic for feature engineering

---

# 1. Source of Truth: VacancyGuide2026 Building Violation Tiers

The VacancyGuide2026.pdf (received from Taylor Popp, Feb 5 2026) defines three 
violation tiers for the **building model** on pages 6–7. Each tier is specified as 
a SQL `IN (...)` clause with an exact list of violation codes.

Key rules from the guide (p.7, footnote 2):

- Only violations with `violationstatus = 'OPEN'` count
- Only violations that are `"current"` count (i.e., the inspection occurred more 
  recently than the most recent construction or demolition)
- **Only the highest applicable tier is scored per property** — not cumulative across tiers

We now transcribe these code lists verbatim from the PDF.

```{r vacancyguide-codes}
# ============================================================
# VacancyGuide2026.pdf, page 6–7: VACANT BUILDING INDICATORS
# Transcribed directly from the SQL IN() clauses
# ============================================================

# Level 1: Add one (1) to Vacant Building Score
# Source: VacancyGuide p.6, first L+I Violations bullet
guide_level1 <- c(
  "PM15-304.10",   # EXTERIOR STRUCT STAIRS, ETC
  "PM15-304.11",   # EXTERIOR STRUCT CHIM TOWERS
  "PM15-304.19",   # DOOR AND WINDOW OCCUPIED/MIXED
  "PM15-304.19O",  # DOOR AND WINDOW OCCUPIED/MIXED
  "PM15-304.2",    # EXTERIOR STRUCT PROTECTIVE TRE
  "PM15-304.8",    # EXTERIOR STRUCT DEC FEATURES
  "PM15-308.1",    # RUBBISH AND GARBAGE
  "PM15-309.1",    # PEST ELIMINATION INFESTATION
  "PM15-309.3",    # PEST ELIM INFEST SGLE OCCUP
  "PM15-309.4",    # PEST ELIMINATION INFESTATION
  "PM-302.2/1",    # EXT A-CLEAN/EXTERMINATE
  "PM-302.2/2",    # EXT A-CLEAN RUBBISH/GARBAGE
  "PM-302.3/1",    # EXT A-CLEAN WEEDS/PLANTS
  "PM-302.4/6",    # DRAINAGE-DOWNSPOUT CONNECT
  "PM-302.4/7",    # DRAINAGE-DOWNSPOUT REPR/REPLC
  "PM-304.1/1",    # EXT S-EXTERIOR REPAIR/MAINTAIN
  "PM-304.6/1",    # EXT S-CHIMMEYS/TOWERS MAINTAIN
  "PM-304.8/1",    # EXT S-WINDOW REPAIR/MAINTAIN
  "PM-304.8/15",   # EXT S-DOOR WEATHERTIGHT
  "PM-304.8/16",   # EXT S-WINDOW REPAIR/REPLACE
  "PM-304.8/17",   # EXT S-DOOR REPAIR/REPLACE
  "PM-304.8/18",   # EXT S-DOORFRAME REPAIR/REPLACE
  "PM-304.8/19",   # EXT S-WINDOW/FRAME RODENTPROOF
  "PM-304.8/2",    # EXT S-DOORS MAINTAIN
  "PM-304.8/20",   # EXT S-DOORS/FRAMES RODENTPROOF
  "PM-304.8/21"    # MIX USE PROP REPLAC W/D 80%
)

# Level 2: Add two (2) to Vacant Building Score
# Source: VacancyGuide p.7, second L+I Violations bullet
guide_level2 <- c(
  "PM15-110.7",    # ID WORK SCHEDULE REQ
  "PM15-302.1",    # EXTERIOR AREA SANITATION
  "PM15-302.4",    # EXTERIOR AREA WEEDS
  "PM15-302.5",    # EXTERIOR AREA RODENT HARBORAGE
  "PM15-304.13",   # EXTERIOR STRUCTURE WINDOWS
  "PM15-304.15",   # EXTERIOR STRUCTURE DOORS
  "PM15-304.16",   # EXTERIOR STRUCTURE HATCHWAY
  "PM15-304.4",    # EXTERIOR STRUCT STRUCTURAL MEM
  "PM15-304.5",    # EXTERIOR STRUCTURE FOUNDATIONS
  "PM15-304.6",    # EXTERIOR STRUCTURE WALLS
  "PM15-304.7",    # EXTERIOR STRUCT ROOF DRAINAGE
  "PM-304.2/1",    # EXT S-FOUNDATION WALLS, MAINTA
  "PM-304.3/1",    # EXT S-WALLS REPAIR/MAINTAIN
  "PM-304.3/2",    # EXT S-WALLS MAKE WEATHERTIGHT
  "PM-304.3/3",    # EXT S-WALLS REPAIR BULGING
  "PM-304.3/4",    # EXT S-WALLS REPAIR CRACKS
  "PM-304.3/5",    # EXT S-WALLS REPAIR STUCCO
  "PM-304.5/2",    # EXT S-EXPOSED SURFACES MAINTAI
  "PM-304.5/3"     # EXT S-CORNICE BARGEBOARD DEFEC
)

# Level 3: Add three (3) to Vacant Building Score
# Source: VacancyGuide p.7, third L+I Violations bullet
guide_level3 <- c(
  "9-3905",        # VACANT STRUCTURE LICENSE
  "CP-103",        # LICENSE - VACANT BUILDING
  "PM-102.4/2",    # LICENSE-VAC RES BLDG
  "PM-102.4/4",    # LICENSE-VAC COMM BLDG
  "PM15-108.1",    # UNSAFE STRUCTURE
  "PM15-108.2",    # VACANT AND OPEN
  "PM15-109.1",    # UNFIT STRUCTURE
  "PM15-110.1",    # ID STRUCTURE
  "PM15-301",      # VACANT STRUCTURE AND LAND
  "PM15-304.1",    # EXTERIOR STRUCTURALLY UNSAFE
  "PM15-304.19V",  # DOOR AND WINDOW VACANT
  "PM15-304.1A",   # EXTERIOR STRUCT UNSAFE COND 1
  "PM15-304.1B",   # EXTERIOR STRUCT UNSAFE COND 2
  "PM15-304.1C",   # EXTERIOR STRUCT UNSAFE COND 3
  "PM15-304.1D",   # EXTERIOR STRUCT UNSAFE COND 4
  "PM15-304.1E",   # EXTERIOR STRUCT UNSAFE COND 5
  "PM15-304.1F",   # EXTERIOR STRUCT UNSAFE COND 6
  "PM15-304.1G",   # EXTERIOR STRUCT UNSAFE COND 7
  "PM15-304.1H",   # EXTERIOR STRUCT UNSAFE COND 8
  "PM15-304.1I",   # EXTERIOR STRUCT UNSAFE COND 9
  "PM15-304.1J",   # EXTERIOR STRUCT UNSAFE COND 10
  "PM15-901.1",    # VACANT PROPERTIES-GENERAL
  "PM15-901.2",    # VACANT PROP-BLIGHTED COND
  "PM15-902.1",    # LARGE VACANT PROP-OPENINGS
  "PM-304.4/1",    # EXT S-ROOF REPAIR
  "PM-306.0/1",    # VACANT PROP STANDARD
  "PM-306.0/90",   # VACANT BLDG KEEP CLEAN GET LIC
  "PM-306.2/1",    # VACANT BLDG UNSECURED COUNT
  "PM-306.2/4",    # VAC PROP REPLAC WIN/DRS 80%
  "PM-306.2/90",   # VACANT BLDG - VOTA
  "PM-306.2/91",   # CP VAC PROP REPLAC WIN/DRS 80%
  "PM-306.2/92",   # VAC PROP REPLAC WIN/DRS 80% CP
  "PM-307.1/1",    # VACANT PROP UNSAFE
  "PM-307.1/11",   # WALL DETERIORATION
  "PM-307.1/12",   # WALL PARTIALLY COLLAPSED
  "PM-307.1/13",   # WALL FIRE DAMAGED
  "PM-307.1/19",   # VACANT BLDG UNSEC/UNSAFE CSU
  "PM-307.1/2",    # ROOF DETERIORATION
  "PM-307.1/3",    # ROOF PARTIALLY COLLAPSED
  "PM-307.1/4",    # ROOF FIRE DAMAGED
  "PM-307.1/5",    # F/C DETERIORATED
  "PM-307.1/6",    # FLOOR/CEILING PARTIALLY COLL
  "PM-307.1/7",    # FLOOR/CEILING FIRE DAMAGED
  "PM-307.1/8",    # WALL BULGED
  "PM-307.1/9",    # WALL FRACTURED
  "PM-307.1/90",   # VACANT BLDG - UNSAFE
  "PM-307.2/1",    # VACANT PROP UNSAFE EQUIPMENT
  "PM-307.3/1",    # UNFIT PROP STANDARD
  "PM-307.3/2",    # UNFIT PROP CEASED
  "PM-307.3/90",   # UNSAFE - UNFIT NO EVAC
  "PM-307.3/91",   # UNSAFE - UNFIT RELOCATE
  "PM-308.1/1",    # ROOF DETERIORATED ID
  "PM-308.1/10",   # WALL DETERIORATED IDF
  "PM-308.1/11",   # WALL PARTIALLY COLLAPSED ID
  "PM-308.1/12",   # WALL FIRE DAMAGED ID
  "PM-308.1/2",    # ROOF PARTIALLY COLLAPSED ID
  "PM-308.1/3",    # ROOF FIRE DAMAGED ID
  "PM-308.1/4",    # F/C DETERIORATED ID COLLAPSE
  "PM-308.1/5",    # FLOOR/CEILING PARTIALLY C ID
  "PM-308.1/6",    # FLOOR/CEILING FIRE DAMAGED ID
  "PM-308.1/7",    # WALL BULGED ID
  "PM-308.1/8",    # WALL CRACK/FRACT ID
  "PM-308.1/9"     # WALL LOOSE/MISS BRICKS ID
)

# Also from VacancyGuide p.5: VACANT LAND violation codes
# These are separate from the building model
guide_land <- c(
  "CP-102",        # LICENSE - VACANT LOT
  "PM-102.4/1",    # LICENSE-VAC LOT
  "PM-102.4/3",    # LICENSE-VAC WHARF/PIER
  "PM-302.2/4",    # EXT A-VACANT LOT CLEAN/MAINTAI
  "PM-304.9/2",    # PIER-VACANT FAULTY/NO BARRIER
  "PM-304.9/3",    # PIER-VACANT POST SIGNS
  "PM-304.9/4",    # PIER-VACANT OBTAIN LICENSE
  "PM-306.0/2",    # VACANT LOT STANDARD
  "PM-306.0/91"    # VACANT LOT KEEP CLEAN GET LIC
)

cat("=== VacancyGuide Building Code Counts ===\n")
cat("Level 1 (+1 point):", length(guide_level1), "codes\n")
cat("Level 2 (+2 points):", length(guide_level2), "codes\n")
cat("Level 3 (+3 points):", length(guide_level3), "codes\n")
cat("Building total:     ", length(guide_level1) + length(guide_level2) + length(guide_level3), "codes\n")
cat("\nLand violations:    ", length(guide_land), "codes\n")
cat("Grand total:        ", length(guide_level1) + length(guide_level2) + length(guide_level3) + length(guide_land), "codes\n")

# Verify no duplicates across building tiers
building_all <- c(guide_level1, guide_level2, guide_level3)
stopifnot("Duplicate codes across building tiers!" = length(building_all) == length(unique(building_all)))
cat("\n✓ No duplicate codes across building tiers\n")

# Check overlap between building and land
bldg_land_overlap <- intersect(building_all, guide_land)
cat("Building-Land overlap:", length(bldg_land_overlap), "codes\n")
if (length(bldg_land_overlap) > 0) cat("  ", paste(bldg_land_overlap, collapse = ", "), "\n")
```

**Key observation:** The VacancyGuide uses **`r length(building_all)` codes** 
for the building model alone. These are not all "vacancy" codes — many are general 
building maintenance codes (rubbish, windows, drainage) that the city's model treats 
as indirect vacancy signals.

---

# 2. Our Previous Approach: The 54 "Vacancy" Codes

In `00_data_inventory.Rmd`, we identified 54 violation codes from 
`VIOLATION_DEFINITION.csv` where the title or definition contains the word "vacant."

```{r our-54-codes}
violation_def <- read_csv("../rawdata/VIOLATION_DEFINITION.csv", show_col_types = FALSE) %>%
  janitor::clean_names()

our_54 <- violation_def %>%
  filter(str_detect(tolower(violation_code_title), "vacant") |
         str_detect(tolower(violation_definition), "vacant")) %>%
  pull(violation_code)

cat("Codes found by searching 'vacant' in title/definition:", length(our_54), "\n")
```

---

# 3. The Gap: Cross-Comparison

This is the critical analysis. How much overlap is there between the VacancyGuide's 
108 building codes and our keyword-based 54 codes?

```{r cross-comparison}
overlap     <- intersect(building_all, our_54)
in_guide_not_54 <- setdiff(building_all, our_54)
in_54_not_guide <- setdiff(our_54, building_all)

cat("=== Cross-Comparison ===\n\n")
cat("In BOTH VacancyGuide building tiers AND our 54:", length(overlap), "\n")
cat("In VacancyGuide but NOT in our 54:             ", length(in_guide_not_54), "\n")
cat("In our 54 but NOT in VacancyGuide building:    ", length(in_54_not_guide), "\n")
```

### 3a. Codes in VacancyGuide but Missing from Our 54

These are the codes we would have **missed entirely** if we only used our 54-code list 
for feature engineering. They are mostly general building maintenance codes — 
the word "vacant" does not appear in their title or definition.

```{r guide-only-codes}
# For each missing code, show which tier it belongs to and its title
guide_only_df <- tibble(code = in_guide_not_54) %>%
  mutate(
    tier = case_when(
      code %in% guide_level1 ~ "Level 1 (+1)",
      code %in% guide_level2 ~ "Level 2 (+2)",
      code %in% guide_level3 ~ "Level 3 (+3)"
    )
  ) %>%
  left_join(
    violation_def %>% select(violation_code, violation_code_title),
    by = c("code" = "violation_code")
  ) %>%
  arrange(tier, code)

guide_only_df %>%
  kable(col.names = c("Code", "VacancyGuide Tier", "Title (from VIOLATION_DEFINITION)"),
        caption = paste0(length(in_guide_not_54),
                         " codes in VacancyGuide building tiers but NOT in our 54 'vacancy' codes")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 12) %>%
  row_spec(which(guide_only_df$tier == "Level 1 (+1)"), background = "#FFF9C4") %>%
  row_spec(which(guide_only_df$tier == "Level 2 (+2)"), background = "#FFE0B2") %>%
  row_spec(which(guide_only_df$tier == "Level 3 (+3)"), background = "#FFCDD2")
```

**Why this gap exists:** Our 54-code list was built by text-searching for "vacant" in 
violation titles/definitions. But VacancyGuide Level 1 and Level 2 are almost entirely 
made up of general property maintenance codes (rubbish, drainage, window repair, pest 
control, etc.) — they don't contain the word "vacant" because they can occur at any 
property. The city's model includes them because L&I observed that these violations 
*tend to cluster on properties trending toward vacancy*, even though they also occur 
at occupied properties.

### 3b. Our 54 Codes Not in VacancyGuide Building Tiers

These codes contain "vacant" in their title but are NOT used in the VacancyGuide 
building model. Why?

```{r our-only-codes}
our_only_df <- tibble(code = in_54_not_guide) %>%
  left_join(
    violation_def %>% select(violation_code, violation_code_title),
    by = c("code" = "violation_code")
  ) %>%
  mutate(
    likely_reason = case_when(
      code %in% guide_land ~ "Used in LAND model, not building",
      str_detect(tolower(violation_code_title), "lot") ~ "Land/lot-specific code",
      str_detect(code, "^F-|^FC-") ~ "Fire safety code — not in VacancyGuide scoring",
      str_detect(code, "9-3904") ~ "Vacant LOT license (land model)",
      str_detect(code, "9-3906|9-3907") ~ "Waterfront/managing agent — not in scoring",
      str_detect(code, "PM15-902") ~ "Large vacant property sub-codes (902.3, 902.7 not scored)",
      TRUE ~ "Review needed"
    )
  ) %>%
  arrange(code)

our_only_df %>%
  kable(col.names = c("Code", "Title", "Likely Reason Not in Building Tiers"),
        caption = paste0(length(in_54_not_guide),
                         " codes in our 54 but NOT in VacancyGuide building tiers")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 12)
```

### 3c. Overlap — Codes in Both Sets

```{r overlap-codes}
overlap_df <- tibble(code = overlap) %>%
  mutate(
    tier = case_when(
      code %in% guide_level1 ~ "Level 1 (+1)",
      code %in% guide_level2 ~ "Level 2 (+2)",
      code %in% guide_level3 ~ "Level 3 (+3)"
    )
  ) %>%
  left_join(
    violation_def %>% select(violation_code, violation_code_title),
    by = c("code" = "violation_code")
  ) %>%
  arrange(tier, code)

overlap_df %>%
  kable(col.names = c("Code", "VacancyGuide Tier", "Title"),
        caption = paste0(length(overlap),
                         " codes that appear in BOTH our 54 and VacancyGuide building tiers")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 12)
```

**Notice:** All `r length(overlap)` overlapping codes are in **Level 3** — the most 
severe tier. This makes sense: Level 3 contains codes that explicitly reference 
"vacant," "unsafe," or "unfit" in their titles. Levels 1 and 2 are composed entirely 
of general maintenance codes that don't mention "vacant."

### 3d. Summary Diagram

```{r summary-counts, fig.width=8, fig.height=4}
summary_df <- tribble(
  ~category, ~count, ~group,
  "Level 1 in Guide, NOT in 54",  sum(in_guide_not_54 %in% guide_level1), "Guide only",
  "Level 2 in Guide, NOT in 54",  sum(in_guide_not_54 %in% guide_level2), "Guide only",
  "Level 3 in Guide, NOT in 54",  sum(in_guide_not_54 %in% guide_level3), "Guide only",
  "Level 3 overlap (in both)",    length(overlap), "Overlap",
  "In 54 only (not in Guide bldg)", length(in_54_not_guide), "54 only"
)

summary_df %>%
  mutate(category = fct_inorder(category)) %>%
  ggplot(aes(x = category, y = count, fill = group)) +
  geom_col() +
  geom_text(aes(label = count), vjust = -0.3, size = 4) +
  scale_fill_manual(values = c("Guide only" = "#E53935", "Overlap" = "#43A047", "54 only" = "#1E88E5")) +
  labs(title = "VacancyGuide Building Codes vs. Our 54 'Vacancy' Codes",
       subtitle = "85 codes in the Guide are NOT in our 54 — mostly Level 1 & 2 maintenance codes",
       x = NULL, y = "Number of Codes", fill = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1, size = 10))
```

---

# 4. Correct Complete Tier Mapping

Now we build the **definitive reference table** that downstream feature engineering 
code will use. Every code gets its tier assignment directly from VacancyGuide.

```{r complete-mapping}
# Build the complete building tier mapping
building_tier_map <- bind_rows(
  tibble(code = guide_level1, guide_tier = "Level 1", guide_weight = 1L),
  tibble(code = guide_level2, guide_tier = "Level 2", guide_weight = 2L),
  tibble(code = guide_level3, guide_tier = "Level 3", guide_weight = 3L)
) %>%
  left_join(
    violation_def %>% select(violation_code, violation_code_title),
    by = c("code" = "violation_code")
  ) %>%
  mutate(
    in_our_54   = code %in% our_54,
    in_ovs_def  = code %in% c("9-3904", "9-3905", "PM15-901.1", "PM15-901.2", "PM15-301",
                                "CP-102", "CP-103", "PM-102.4/1", "PM-102.4/2", 
                                "PM-102.4/3", "PM-102.4/4"),
    code_type = case_when(
      in_our_54 & in_ovs_def ~ "Vacancy code + OVS source",
      in_our_54              ~ "Vacancy code (not in OVS)",
      TRUE                   ~ "General maintenance code"
    )
  ) %>%
  arrange(guide_weight, code)

cat("=== Complete Building Tier Mapping ===\n")
cat("Total codes:", nrow(building_tier_map), "\n\n")

building_tier_map %>%
  count(guide_tier, code_type) %>%
  pivot_wider(names_from = code_type, values_from = n, values_fill = 0) %>%
  kable(caption = "Breakdown by tier and code type") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### 4a. Full Level 1 Codes (+1 point)

```{r level1-table}
building_tier_map %>%
  filter(guide_weight == 1) %>%
  select(code, violation_code_title, code_type) %>%
  kable(col.names = c("Code", "Title", "Type"),
        caption = paste0("Level 1: ", sum(building_tier_map$guide_weight == 1),
                         " codes — Minor deterioration, may occur at occupied properties")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 12)
```

### 4b. Full Level 2 Codes (+2 points)

```{r level2-table}
building_tier_map %>%
  filter(guide_weight == 2) %>%
  select(code, violation_code_title, code_type) %>%
  kable(col.names = c("Code", "Title", "Type"),
        caption = paste0("Level 2: ", sum(building_tier_map$guide_weight == 2),
                         " codes — Moderate structural damage, stronger vacancy signal")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 12)
```

### 4c. Full Level 3 Codes (+3 points)

```{r level3-table}
building_tier_map %>%
  filter(guide_weight == 3) %>%
  select(code, violation_code_title, code_type) %>%
  kable(col.names = c("Code", "Title", "Type"),
        caption = paste0("Level 3: ", sum(building_tier_map$guide_weight == 3),
                         " codes — Severe: vacant, unsafe, unfit, or major collapse")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 12)
```

---

# 5. Vacant Land Codes (Separate Model)

For completeness, here are the land model violation codes (VacancyGuide p.5).

```{r land-codes}
land_tier_map <- tibble(code = guide_land, guide_weight = 1L) %>%
  left_join(
    violation_def %>% select(violation_code, violation_code_title),
    by = c("code" = "violation_code")
  ) %>%
  arrange(code)

land_tier_map %>%
  kable(col.names = c("Code", "Weight", "Title"),
        caption = paste0("Vacant Land: ", nrow(land_tier_map),
                         " codes — All weighted +1")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 12)
```

---

# 6. Validation Against violations.csv

Now let's check: how many of these codes actually appear in our violation data, 
and with what frequency?

```{r validate-against-data}
violations <- read_csv("../rawdata/violations.csv", show_col_types = FALSE) %>%
  janitor::clean_names()

# Count OPEN violations for each building tier code
building_viol_counts <- violations %>%
  filter(violationcode %in% building_all,
         tolower(violationstatus) == "open") %>%
  count(violationcode, name = "n_open") %>%
  right_join(building_tier_map %>% select(code, guide_tier, guide_weight),
             by = c("violationcode" = "code")) %>%
  replace_na(list(n_open = 0)) %>%
  arrange(guide_weight, desc(n_open))

cat("=== OPEN violations by tier ===\n")
building_viol_counts %>%
  group_by(guide_tier) %>%
  summarise(
    n_codes_with_data = sum(n_open > 0),
    n_codes_total     = n(),
    total_open_viols  = sum(n_open),
    .groups = "drop"
  ) %>%
  kable(col.names = c("Tier", "Codes with ≥1 OPEN", "Total Codes", "Total OPEN Violations"),
        caption = "How many VacancyGuide codes actually have OPEN records in our data?") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

### Top 20 Most Frequent OPEN Violation Codes (Building Tiers)

```{r top-codes}
building_viol_counts %>%
  filter(n_open > 0) %>%
  head(20) %>%
  kable(col.names = c("Code", "OPEN Count", "Tier", "Weight"),
        caption = "Top 20 building tier codes by OPEN violation count") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 12)
```

### Codes with Zero OPEN Violations

```{r zero-codes}
zero_codes <- building_viol_counts %>% filter(n_open == 0)

cat(nrow(zero_codes), "building tier codes have zero OPEN violations in current data.\n")
cat("These may be legacy codes or codes whose violations have all been resolved.\n\n")

zero_codes %>%
  kable(col.names = c("Code", "OPEN Count", "Tier", "Weight")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 12)
```

---

# 7. Feature Engineering Implications

### Methodological concern: avoid importing external bias

The VacancyGuide tier system (Level 1/2/3) was constructed through expert judgment 
and trial-and-error, not statistical analysis (confirmed by Tim Haynes at the Jan 27 
kickoff meeting). Using these predetermined tiers directly as ML features risks:

1. **Importing subjective bias** — the tier assignments reflect someone else's 
   assumptions about which violations matter more, without empirical validation.
2. **Collinearity** — tier-based features are derived from the same underlying 
   violation codes, so `max_violation_tier`, `has_level2`, and `n_violations_L2` 
   will be mechanically correlated.
3. **Information loss** — collapsing 108 codes into 3 tiers discards variation 
   that the ML model could learn from.

Per instructor feedback (Prof. Fichman, Feb 18 2026): feature engineering should 
start from **raw violation data** and use **univariate correlation analysis** to 
determine which features are genuinely predictive, rather than assuming the city's 
groupings are correct.

### Recommended approach: raw features first, tiers as comparison

**Track A (Primary): Raw violation features — let the model learn**

Built directly from violations.csv without imposing any external grouping.

| Feature | Logic |
|---------|-------|
| `n_violations_total` | Count of ALL violations on the property (any code) |
| `n_violations_open` | Count of currently OPEN violations |
| `n_violation_codes_distinct` | Number of distinct violation codes on the property |
| `oldest_open_violation_days` | Days since the oldest OPEN violation |
| `newest_violation_days` | Days since the most recent violation |
| `has_vacancy_code` | Binary: any of the 54 vacancy keyword codes present |
| `has_structural_code` | Binary: any PM15-304.x / PM-307.x structural codes present |
| `has_unsafe_code` | Binary: PM15-108.1 (Unsafe) or PM15-108.2 (Vacant & Open) present |
| `has_fire_code` | Binary: any F-xxx / FC-xxx fire safety codes present |
| `n_vacancy_violations` | Count of violations with vacancy keyword codes |

These features preserve the full variation in the data. The ML model determines 
which violation patterns are most predictive of OVS — we do not pre-impose weights.

**Track B (Comparison): VacancyGuide tier features — for benchmarking**

Uses the 108-code tier mapping from this document. These features are included 
**only to compare against Track A** and to evaluate whether the city's existing 
groupings add predictive value beyond raw features.

| Feature | Logic |
|---------|-------|
| `max_violation_tier` | Highest tier (0/1/2/3) among OPEN violations |
| `guide_score_raw` | Sum of tier weights across OPEN violations (replicates VPI logic) |
| `has_level1_violation` | Binary: any Level 1 OPEN violation present |
| `has_level2_violation` | Binary: any Level 2 OPEN violation present |
| `has_level3_violation` | Binary: any Level 3 OPEN violation present |

**Track C (Validation): Univariate correlation analysis**

Before any modeling, we will:

1. Compute correlation of each Track A and Track B feature with OVS
2. Build a feature-to-feature correlation heatmap to identify collinearity
3. Create facet plots of continuous features split by OVS (0 vs 1)
4. Use these results to decide which features enter the model and whether 
   tier-based groupings are empirically justified

This ensures we are not importing the city's heuristic assumptions as ground truth, 
while still being able to evaluate whether their expert judgment aligns with the data.

---

# 8. Non-Violation Indicators from VacancyGuide

For completeness, here are the other building model indicators from VacancyGuide 
(pages 6–7) that are NOT violation-based:

```{r non-violation-indicators}
tribble(
  ~indicator, ~source, ~weight, ~data_status,
  "Water: NB9 (Vacant) installation status, excl. type 24", "Water Revenue", "+1", "Pending legal",
  "Water: No usage 4+ months (excl. NB3/NB9/NB2/NB8/type24/fire/commercial/stormwater)", "Water Revenue", "+1", "Pending legal",
  "PWD: Side yard exemption for stormwater billing", "PWD", "Reset to 0", "Pending legal",
  "Permit: completed in last 4 years (plumbing/demolition/electrical/fire suppression/mechanical/building new construction/alteration)", "L+I Permits", "-1", "To download",
  "License: Active rental license (revenuecode = 3202)", "Business Licenses", "-1", "Available",
  "License: Active vacancy license (revenuecode IN 3219, 3634)", "Business Licenses", "+2", "Available",
  "Permit: Completed demolition without subsequent new construction", "L+I Permits", "Reset to 0", "To download"
) %>%
  kable(col.names = c("Indicator", "Source", "Weight", "Data Status"),
        caption = "Non-violation building indicators from VacancyGuide") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE, font_size = 12)
```

---

# 9. Corrections to Previous Work

This analysis identified the following errors in our earlier `violation_tier_mapping.xlsx`:

1. **VacancyGuide Building Tiers sheet listed 11 codes; the actual count is 108.** 
   The Excel showed 3 Level 1, 3 Level 2, and 5 Level 3 codes as representative 
   examples, but these were incomplete. The correct counts are 26 / 19 / 63.

2. **"54 vacancy codes" ≠ VacancyGuide tier codes.** The 54-code list was built by 
   keyword search ("vacant") and captures a different dimension. Only 23 codes overlap, 
   all in Level 3. The entire Level 1 (26 codes) and Level 2 (19 codes) were absent 
   from our 54.

3. **Feature Plan assumed two tracks; actually needs three.** The original plan 
   correctly noted that Level 1/2 codes are "not in our 54" but underestimated the 
   scope (6 codes mentioned vs. 85 actual). The feature plan is still valid in 
   structure but needs to use the complete code lists from this document.

---

# 10. Export Reference Tables

Save the tier mapping as CSV for use in downstream Rmd files.

```{r export}
# Ensure output directory exists
dir.create("../data", showWarnings = FALSE, recursive = TRUE)

# Building tier mapping
building_tier_map %>%
  select(code, guide_tier, guide_weight, violation_code_title, in_our_54, in_ovs_def) %>%
  write_csv("../data/building_tier_mapping.csv")

# Land codes
land_tier_map %>%
  write_csv("../data/land_violation_codes.csv")

# Our 54 vacancy codes (still useful as a separate reference)
tibble(code = our_54) %>%
  left_join(violation_def %>% select(violation_code, violation_code_title),
            by = c("code" = "violation_code")) %>%
  write_csv("../data/vacancy_keyword_codes_54.csv")

cat("Exported:\n")
cat("  ../data/building_tier_mapping.csv (", nrow(building_tier_map), " rows)\n")
cat("  ../data/land_violation_codes.csv  (", nrow(land_tier_map), " rows)\n")
cat("  ../data/vacancy_keyword_codes_54.csv (", length(our_54), " rows)\n")
```

---

# References

- **VacancyGuide2026.pdf** — "Vacant Property Indicators Model," updated 2/3/2026, 
  received from Taylor Popp via email on Feb 5, 2026. Pages 6–7 contain the building 
  model SQL queries; page 5 contains the land model queries; pages 8–10 contain the 
  reference violation code table.
- **VIOLATION_DEFINITION.csv** — L+I violation code lookup table, downloaded from 
  OpenDataPhilly (L+I Code Violations attachment).
- **Kickoff meeting notes** (Jan 27, 2026) — Tim Haynes confirmed: "only the most 
  severe violation is scored per property"; weights were set through trial and error.
